Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-24 BMC-Response-Round1.tex
Updates from ShareLaTeX

\Comment{
Moreover, I cannot find a definition of several variables. (e.g., What is $S$, $S_i$, $x$, $xp$, $xd$, $vp$, $bu{chi}$ (P9L24) ? Where "its dependence dep $f$" comes from?)
Because this manuscript contains many variables and abbreviations, it is recommended to have a table or image to explain the variables and their actual examples following the sampling example in Fig1. The description of the brief aim and summary of each step would be a good guide for readers.}%
\Answer{We included additional explanations and took care to define these variables. We considered the incl further advantages.}

>>>>

\Comment{
Moreover, I cannot find a definition of several variables. (e.g., What is $S$, $S_i$, $x$, $xp$, $xd$, $vp$, $bu{chi}$ (P9L24) ? Where "its dependence dep $f$" comes from?)
Because this manuscript contains many variables and abbreviations, it is recommended to have a table or image to explain the variables and their actual examples following the sampling example in Fig1. The description of the brief aim and summary of each step would be a good guide for readers.}%
\Answer{We included additional explanations and took care to define all variables (like the sequence $S$). We considered the inclusion of a table, but this seems to offer little further advantages.}

### 2019-01-24 RNARedPrint.tex
Updates from ShareLaTeX

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasiblity of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\parHead{Specificities and similarities of multi-target design.}
The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
%
Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
%
Recently, the approach \Software{RNAdesign}~\citep{HoenerzuSiederdissen2013}, coupled with local
search in \RNAblueprint{}~\citep{Hammer2017}, solved the
problem of sampling seeds from the \emph{uniform} distribution for multiple
target structures. \Software{RNAdesign} adopts a graph coloring perspective,
\revised{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform \Def{stochastic backtracking}, inspired by
Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
choosing individual nucleotide assignments with probabilities derived
from the subsolution counts. While, due to its decomposition strategy, 
\Software{RNAdesign} performs much better than the theoretical bound of $O(4^n)$, no attempts were made to characterize or justify its---still exponential---complexity; leaving important theoretical questions of the complexity of counting and uniform sampling open.
As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility. \revised{Substantial improvements of multi-target sampling thus require a systematically redesigned appraoch. To enable a fundamentally broader range of applications in extensions of the sampling method, we build our approach, from the start, on established concepts in computer science.}

@@

%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$. Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all strucutures) from the Boltzmann distribution defined by} 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
%
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
  \end{equation}
  for specfic weights $\pi_1,\dots,\pi_k$.

\subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}. Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that

@@

The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each dependencing only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$

@@

\smallskip
\noindent
The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
$\min_{u\in T} |\chi(u)| - 1.$ 
%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
\end{definition}
%
\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependend subproblems to ensure consistency in a recursive computation. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}

@@

Since our algorithms iterate over specific sets of sequence positions, we moreover define
the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y=\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
i.e.~for $\Y=\{y_1,\dots,y_{q}\}$,
\begin{align*}
\partseqs(\Y) = \{ & \{ y_i\mapsto v_i \mid i=1,\dots,q \} \\& | (v_1,\dots,v_{q}) \in \{\Ab,\Cb,\Gb,\Ub\}^{q} \}.
\end{align*}
%
We assume the following properties of the given cluster tree (reflecting $\F$):
\begin{itemize}
\item
$T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$;
\item
all edges in the tree decompostion are oriented towards this root;
\item
all sets $\difference{u}{v}$ are singleton: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.
\end{itemize}

@@

\paragraph{Approximating Turner Energy using Simpler Energy Models.}
To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogenous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

\subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and dinucleotide content~\citep{Zhang2013} of sampled RNA sequences. Here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).

@@

\subsection*{Effectively targeting Turner energies using multidimensionnal sampling}
\revised{
We used our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
used multi-dimensional Boltzmann sampling to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp.\ Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in subfigure B). The good
    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogenous weights $1$ and
    $e^\beta$, respectively.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}

@@

We considered a benchmark of six sets of target structures described in \citep{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
%%For each instance of the \Software{Modena}
%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
%Our procedure is tailored to produce sequences with similar Turner
%energy that favor the stability of the target structures having
%moderate \GCb\%; all these properties are desirable objectives
%for RNA design.
Using \revised{multi-dimensional Boltzmann sampling based on} \ourprog{}, for
each instance of the benchmark, we generated at least 1\,000 seed sequences with similar energies for all target structures. 
For this purpose, we determine good common target energies that
can be successfully targeted for all single target structures simultaneously.
Generally, we targeted 60\% \GCb\%.
\revised{As described in Section~\ref{sec:multiBoltzmann}, to find sequences fulfilling our energy and \GCb\% criteria, we iteratively sampled 10\,000 sequences and kept admissible ones while optimizing the weights. In practice, this sampling operation performs quite fast (see \ref{appsec:run-times}) and only little iterations were needed for most of the benchmark inputs as reported in  \ref{appsec:effectivity-seed-generation}.}
We compared the
\Obj{} value of \revised{the derived} sequences against that of seed sequences,
uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
\Obj{} function. At each step, the local search resamples (uniformly
at random) the positions of a randomly selected component in the base
pair dependency graph, accepting the modification only if it results
in a gain. We performed 500 greedy descent steps in the case of
pseudoknot-free datasets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
  LE80}.

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every dataset,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is observed for both terms in \Obj{} (see
Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
produces sequences whose targets are both, substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

@@

\revised{Based on a general framework and efficient algorithms, we introduced a novel approach to design RNA
sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target strucutures and the \GCb-content.
Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models. \revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework. By conducting a proof-of-concept study 
  on an established benchmark set for negative multi-target RNA design (including pseudoknotted instances), we showcase a typical application of our tool \ourprog{}. In this study, 
  the approach generated significantly better seed sequences than
the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
}
%
To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.

>>>>

\begin{abstractbox}
\begin{abstract}
  \parttitle{Background}
  The design of multi-stable RNA molecules has important applications in biology, medicine, and biotechnology. Synthetic design approaches profit strongly from effective in-silico methods, \revised{which substantially reduce the need for costly wet-lab experiments.} \\
%
  \parttitle{Results} We devise a novel approach to a central ingredient of most in-silico
  design methods: the generation of sequences \revised{that fold well into multiple target structures.} 
  % 
  \revised{Based on constraint networks, our approach \ourprog{}} supports generic
  Boltzmann-weighted sampling, which enables the positive design of RNA sequences
  with specific free energies (for each of multiple, possibly pseudoknotted, target structures) and \GCb-content.
%
  Moreover, we study general properties of our approach empirically and generate
  biologically relevant multi-target Boltzmann-weighted designs for a
  \revised{established} design benchmark. \revised{Our results demonstrate the effectivity and feasibility of the method in practice
  as well as the benefits of Boltzmann sampling over the previously best multi-target sampling strategy---even for the case of negative design of multi-stable RNAs.}\\
%
  \revised{Besides empirically studies, we finally justify the algorithmic details due to a fundamental theoretic result about multi-stable RNA design, namely the \#P-hardness of the counting of designs.}
%
  \parttitle{Conclusion} \ourprog{} introduces a novel, flexible and---as we demonstrate---effective approach to multi-target RNA design, which promises broad applicability and extensibility.

@@

\parHead{Specificities and similarities of multi-target design.}
The available methods for multi-target RNA design~\citep{Lyngsoe2012,HoenerzuSiederdissen2013,Taneda2015,Hammer2017} all follow the same overall generate-and-optimize strategy.
%
Faced with the complex constraints due to the multiple targets, early methods such as \Software{Frnakenstein}~\citep{Lyngsoe2012} and \Software{Modena}~\citep{Taneda2015} \revised{do not even attempt to sample sequences systematically from a controlled distribution}, but rely on ad-hoc generation strategies.
%
Recently, the approach \Software{RNAdesign}~\citep{HoenerzuSiederdissen2013}, coupled with local
search in \RNAblueprint{}~\citep{Hammer2017}, solved the
problem of sampling seeds from the \emph{uniform} distribution for multiple
target structures. \Software{RNAdesign} adopts a graph coloring perspective,
\revised{assigning nucleotide symbols (like ``colors'') to the sequence positions, such that compatible nucleotides are assigned to the ends of each base pair.} Initially, the method decomposes the graph hierarchically and then \Def{precomputes} the number of valid sequences
within each subgraph. The decomposition is then reinterpreted as a
decision tree to perform \Def{stochastic backtracking}, inspired by
Ding and Lawrence~\citep{Ding2003}. Uniform sampling is achieved by
choosing individual nucleotide assignments with probabilities derived
from the subsolution counts. While, due to its decomposition strategy, 
\Software{RNAdesign} performs much better than the theoretical bound of $O(4^n)$, no attempts were made to characterize or justify its---still exponential---complexity; leaving important theoretical questions of the complexity of counting and uniform sampling open.
As well, the \Software{RNAdesign}/\RNAblueprint{} approach is specialized to uniform sampling, which limits its direct extensibility. \revised{Substantial improvements of multi-target sampling thus require a systematically redesigned approach. To enable a fundamentally broader range of applications in extensions of the sampling method, we build our approach, from the start, on established concepts in computer science.}

@@

%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$. Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all structures) from the Boltzmann distribution defined by} 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
%
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
  \end{equation}
  for specific weights $\pi_1,\dots,\pi_k$.

\subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}. Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, namely its \emph{dependencies $\dep(f)$}, such that

@@

The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each depending only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$

@@

\smallskip
\noindent
The \Def{treewidth} of a tree decomposition $(T,\chi)$ is defined as
$\min_{u\in T} |\chi(u)| - 1.$ 
%The \Def{treewidth} of a graph $G$ is the smallest treewidth of any tree decomposition of $G$.
\end{definition}
%
\revised{Intuitively, a tree decomposition of an (hyper)graph $G$ is a tree that captures all the vertices and (hyper)edges of $G$, and properly relates dependent sub-problems to ensure consistency in a recursive computation. Figure~\ref{fig:treedec} shows an optimal tree decomposition for a pair of structures under the stacking energy model.}

@@

Since our algorithms iterate over specific sets of sequence positions, we moreover define
the \Def{set $\partseqs(\Y)$ of all partial sequences determining the positions of $\Y=\subseteq\{1,\dots,n\}$} in all combinations of nucleotides $\{\Ab,\Cb,\Gb,\Ub\}$,
i.e.~for $\Y=\{y_1,\dots,y_{q}\}$,
\begin{align*}
\partseqs(\Y) = \{ & \{ y_i\mapsto v_i \mid i=1,\dots,q \} \\& | (v_1,\dots,v_{q}) \in \{\Ab,\Cb,\Gb,\Ub\}^{q} \}.
\end{align*}
%
We assume the following properties of the given cluster tree (reflecting $\F$):
\begin{itemize}
\item
$T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$. If such a root does not exist, it can be added to the tree decomposition and connected to one node in each connected component of $T$;
\item
all edges in the tree decomposition are oriented towards this root;
\item
all sets $\difference{u}{v}$ are singleton: for any given
cluster tree, an equivalent (in term of treewidth) cluster tree can
always be obtained by inserting at most $\Theta(|\X|)$ additional
clusters.
\end{itemize}

@@

\paragraph{Approximating Turner Energy using Simpler Energy Models.}
To capture the realistic Turner model $\EnergyTurner$ more efficiently, we exploit the tight correlation between
$\EnergyTurner$ and the fitted stacking model $\EnergyStacking$
(Supp.\ Mat.~\ref{appsec:modelparameters}). More precisely, we observed a
structure-specific affine dependency between the Turner and stacking
energy models, so that
$\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$
for any structure $R$ and sequence $S$. We inferred the
$(\gamma,\delta)$ parameters from a set of sequences generated with
homogeneous weights $w=e^{\beta}$, tuning only \GCb\% to a
predetermined value.  Finally, we adjusted the targeted energies
within our stacking model to
$\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in
order to reach, on average, the targeted energy
$\EnergyTurner^{\star}$ in the Turner model.

\subsection*{Extension to multidimensional Boltzmann sampling}\label{sec:multiBoltzmann}
The flexibility of our framework allows to support the advanced sampling technique called "multidimensional Boltzmann sampling"~\citep{Bodini2010}, which allows to enforce (probabilistically)  additional, complex properties of the samples through an additional rejection.
This technique was previously used to control \GCb\%~\citep{Waldispuehl2011,Reinharz2013} and di-nucleotide content~\citep{Zhang2013} of sampled RNA sequences. Here, in addition to controlling \GCb\% (our feature $F_0$) we use it to target the free energies $(\TargetE_1,\ldots,\TargetE_k)$ of the individual target structures (features $F_1,\dots,F_m$).

@@

\subsection*{Effectively targeting Turner energies using multi-dimensional sampling}
\revised{
We used our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for
given target structures and weights $\pi_1,\dots,\pi_k$.  Moreover, we
used multi-dimensional Boltzmann sampling to target specific energies and
\GCb\%.  Our tool \ourprog{} evaluates energies according to the
stacking energy model $\EnergyStacking$, whose parameters were fitted
to best approximate Turner energies. As well, we implemented and
fitted a base pair energy model for \ourprog{}, which was not studied
for its targeting performance (both models:
Supp.\ Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or a maximum amount of iterations is reached, the sequences in $P$ are returned.
%

@@

%To reduce the cost of the precomputation, we used a stacking energy model
%as a proxy for the Turner model. Indeed, we observed a strong linear
%dependency between the Turner energy and the
%emphasize that only the simple stacking energies are directly targeted
%(at $\pm$10\% tolerance, except for a few hard instances). Due to our
%linear adjustment, we finally achieve well-controlled energies in the
%realistic Turner energy model.
\begin{figure*}[t]
  \begin{center}
    %{\sf \bfseries A}\
    \includegraphicstop[width=0.97\textwidth]{Figs/energy_shift}\hfill
    %{\sf \bfseries B}\ \includegraphicstop[width=0.55\textwidth]{Figs/PKB00211_PKB00239_0_energy_distribution}
  \end{center}
  \caption{%
    Targeting specific energies for pseudoknotted structures using
    multi-dimensional Boltzmann sampling. \textbf{(A)} Linear fits between the
    energies in the stacking model to the realistic pseudoknot energy
    model by Dirks and Pierce (D\&P) for \revised{initially} sampled sequences and both target structures R1 and R2 (shown in B). The good
    match \revised{(respective $R^2$ values of $0.846$ and $0.841$)} enables more efficient targeting of Turner energies based on
    targeting stacking model energies. \textbf{(B)} \revised{Resulting D\&P energy distributions for the two target structures R1 and R2 when aiming for the respective free energies $(-30,-20)$,
    $(-30,-30)$, $(-25,-25)$, and $(-20,-30)$. These demonstrate the effectivity of our adaptive
    multi-dimensional Boltzmann sampling procedure, especially by} comparing the distributions to those of uniform and Boltzmann sampled sequences, with homogeneous weights $1$ and
    $e^\beta$, respectively.
%    ($\delta=0.05, \delta'=0.1, \gamma=1.1, \psi=20, n=1000$)
  }
  \label{fig:energydist-pk}
\end{figure*}

@@

We considered a benchmark of six sets of target structures described in \citep{Taneda2015}: {\sf 2str}, {\sf 3str}, and {\sf 4str} consist of non-pseudoknotted structures, while {\sf PK60}, {\sf PK80}, and {\sf LE80} contain pseudoknotted structures.
%%For each instance of the \Software{Modena}
%benchmark~\citep{Taneda2015}, we generated $1\,000$ seed sequences at these energies.
%Our procedure is tailored to produce sequences with similar Turner
%energy that favor the stability of the target structures having
%moderate \GCb\%; all these properties are desirable objectives
%for RNA design.
\revised{Based on} \ourprog{}, we generated at least 1\,000 seed sequences with similar energies for all target structures, for
each instance of the benchmark. 
For this purpose, we determine good common target energies that
can be successfully targeted for all single target structures simultaneously.
Generally, we targeted 60\% \GCb\%.
\revised{%
Supp.\ Mat.~\ref{appsec:effectivity-seed-generation}, provides detailed results from this iterative procedure, which works similarily to the previously described multi-dimensional Boltzmann sampling. In particular, we observe that most of the benchmark inputs finish within only few iterations, where each iteration requires  little time (confer Supp.\ Mat.~\ref{appsec:run-times}).}

We compared the
\Obj{} value of \revised{the derived} sequences against that of seed sequences,
uniformly sampled using \RNAblueprint{}.  Moreover, for both sets we
used an adaptive greedy walk~\citep{Hammer2017} to \Def{minimize} the
\Obj{} function. At each step, the local search re-samples (uniformly
at random) the positions of a randomly selected component in the base
pair dependency graph, accepting the modification only if it results
in a gain. We performed 500 greedy descent steps in the case of
pseudoknot-free data-sets {\sf 2str}, {\sf 3str}, and {\sf 4str}; and
200 steps for the pseudoknotted ones {\sf PK60}, {\sf PK80}, and {\sf
  LE80}.

The results, shown in Fig.~\ref{fig:benchmark-results}, reveal that
Boltzmann-sampled sequences outperform uniform seeds on every data-set,
leading to average improvements in \Obj{} values ranging from $7.26$
({\sf LE80}) to $16.05$ ({\sf 2str}) units. Remarkably, this
improvement is observed for both terms in \Obj{} (see
Supp. Mat.~\ref{appsec:contributions-multidefect}).  This means that \ourprog{}
produces sequences whose targets are both, substantially closer to
the sequence ensemble/minimal free energy, and more similarly stable across targets.  In fact, for
every sequence in our benchmark, consisting of 332 sets of target
structures, we observed better \Obj{} for Boltzmann sampling than for
uniform sampling (see Supp. Mat.~\ref{appsec:detailed-benchmark-results}).  Notably,
\ourprog{} performs equally well in the presence of pseudoknots;
the difficulty rather lies in the computation of the \Obj{} function, since
free-energy minimization is costly in the presence of
pseudoknots~\citep{Sheikh2012} and good implementations are scarce.

@@

\revised{Based on a general framework and efficient algorithms, we introduced a novel approach to design RNA
sequences while targeting very specific complex properties}. In particular, we describe the targeting of the free energies of multiple target structures and the \GCb-content.
Our method combines an FPT stochastic sampling algorithm
with multi-dimensional Boltzmann sampling over distributions
controlled by expressive RNA energy models. \revised{We demonstrate that the approach, despite of its theoretical hardness, performs well on typical multi-stable RNA design instances in practice, which is a direct consequence of the approximability of Turner energies as well as the systematic algorithmic framework. By conducting a proof-of-concept study 
  on an established benchmark set for negative multi-target RNA design (including pseudoknotted instances), we showcase a typical application of our tool \ourprog{}. In this study, 
  the approach generated significantly better seed sequences than
the previously best available method (uniform sampling). These results strongly suggest that the presented technique for positive design can be highly beneficial in future negative design approaches.
}
%
To substantiate our work additionally, we establish the $\#${\sf P}-hardness of uniform sampling which,
from a complexity-theoretic point of view, motivates the FPT, tree decomposition-based nature of our method.
